Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-28 Paper_Draft2/Draft2.Rmd
Updating processing algorithms sections

Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018]. We use a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we are able to compare our results to the baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while building a model. It assesses how we build learning algorithms, which often prioritize accuracy over fairness. The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. Similarly to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model outputs. Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values. Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box. To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set [@aif360-oct-2018]. Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.

@@

We chose the reweighing as our pre-processing bias mitigation algorithm. After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio. Overall, these fairness metrics show that the pre-processing algorithm reweighing improves the bias in the COMPAS algorithm.

>>>>

Pre-processing refers to mitigating bias within the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. Training data is where bias is most likely to be introduced.  We use the reweighing pre-processing algorithm from AIF360 which assigns weights to the data. “The advantage of this approach is, instead of modifying the labels, it assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset. The weights are based on frequency counts. However as this technique is designed to work only with classifiers that can handle row-level weights, this may limit your modeling options” [@Ronaghan2019AI]. We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.

In-processing mitigates bias in classifiers while building a model. A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021what]. The in-processing technique we use is the prejudice remover algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset. Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

The prejudice remover is a method for reducing indirect prejudice (i.e. how COMPAS is racially biased because it uses proxy variables for race). The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification. The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by Kamishima et. al. This in turn enforces a classifier’s independence from sensitive information (e.g., race).

Our last approach, post-processing bias mitigation, is used after training a model. Post-processing algorithms equalize the outcomes (i.e., predicted classification values) to mitigate bias instead of adjusting the classifier or the training data [@baxter2021AI]. We use calibrated equalized odds, which “optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective” [insert aif360 website citation here]. An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates. Through the calibrated equalized odds method, we want to decrease bias while also maintaining calibration [@pleiss2017fairness]. [Calibration](https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555) refers to improving a model so that the distribution of predicted outcomes is similar to the distribution of observed probability in the training data.

@@

After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio. Overall, these fairness metrics show that the pre-processing algorithm reweighing improves the bias in the COMPAS algorithm.
